Interisland movements and oceanic swimming of woodland caribou in Newfoundland

### 2020-05-11 paper/manuscript.Rmd
minor edits and fig2 caption

Caribou (*Rangifer tarandus*) are exceptional swimmers. Ample evidence exists
that caribou swim in streams, rivers, and lakes during migration [@Leblond2016],
to avoid predators [@Bergerud1985], and to access islands during calving
[@Bergerud1990]. Even for adept swimmers like caribou, that have hollow hair
shafts that aid in flotation, the energetic expenditure of swimming for
quadrupedal mammals is significantly higher than walking or running, and
drowning is also possible [@Miller1985]. Despite the abundance of coastal and
island caribou herds, only a handful of studies have documented caribou swimming
in the ocean, and most of these lack observation, but rather deduce swimming
based on presence of caribou on a previously uninhabitated island  [e.g.
@Miller1995; @Miller2002; @Jeffery2007; @Ricca2012]. Caribou may swim between
islands in the ocean for similar reasons they swim in freshwater [@Leblond2016],
mainly concerning predator avoidance or movement between habitats. We propose
that one such explanation, the forage limitation hypothesis, is the most likely
explanation for oceanic swimming for caribou that live on islands in the Fogo
Island archipelago.

Caribou were introduced to Fogo Island (n = 26 animals between 1964–67) and
Change Island (n = 5 animals in 1964) from Newfoundland as part of a series of
translocations and introductions throughout the province [@Bergerud1989].
Typical caribou predators such as black bears (*Ursus americanus*) and wolves
(*Canis lupus*) are absent from Fogo Island; the dominant predator is coyote
(*Canis latrans*), which predate calves but it remains unclear what threat they
pose, if any, to adult caribou [@Rayl2014]. Given the minimal risk of predation to adults,
we propose that forage limitation is more likely to drive movement between
islands than predator avoidance. For several decades after their introduction to
Fogo Island, the caribou population did not exceed ~100 individuals
[@Bergerud1989, Newfoundland and Labrador Wildlife Division, unpublished data].
During the 1990s, population density reached ~300 animals; anecdotal evidence
suggests caribou were first seen on nearby islands during this same period.
Increased competition for resources on Fogo due to this increase in density
could have precipitated movement to other islands.

@@

Forage limitation may be the ultimate explanation for caribou swimming between
oceanic islands, but a number of proximate mechanisms remain untested. Our
observations of caribou swimming in the ocean to five questions for future
research:

@@

(ref:fig2) A) Map of the Fogo Island archipelago with swimming events between islands. B) Histogram displaying the distribution of swimming events throughout the year. Note, colours correspond to individual caribou. C) Inset of swimming events between small islands on the northern coast of Fogo island. D) Inset of swimming events between islands on the southern coast of Fogo island. Note, the grey diamond represents the location we observed an adult male caribou swimming (see Figure 1).

>>>>

Caribou (*Rangifer tarandus*) are exceptional swimmers. Ample evidence exists
that caribou swim in streams, rivers, and lakes during migration [@Leblond2016],
to avoid predators [@Bergerud1985], and to access islands during calving
[@Bergerud1990]. Even for adept swimmers like caribou, that have hollow hair
shafts that aid in flotation, the energetic expenditure of swimming for
quadrupedal mammals is significantly higher than walking or running, and
drowning is also possible [@Miller1985]. Despite the abundance of coastal and
island caribou herds, only a handful of studies have documented caribou swimming
in the ocean, and most of these lack observation, but rather deduce swimming
based on presence of caribou on a previously uninhabitated island  [e.g.
@Miller1995; @Miller2002; @Jeffery2007; @Ricca2012]. Caribou may swim between
islands in the ocean for similar reasons they swim in freshwater [@Leblond2016],
mainly concerning predator avoidance or to access new habitats. We propose
that one such explanation, the forage limitation hypothesis, is the most likely
explanation for oceanic swimming for caribou that live on islands in the Fogo
Island archipelago.

Caribou were introduced to Fogo Island (n = 26 animals between 1964–67) and
Change Island (n = 5 animals in 1964) from Newfoundland as part of a series of
translocations and introductions throughout the province [@Bergerud1989].
Typical caribou predators such as black bears (*Ursus americanus*) and wolves
(*Canis lupus*) are absent from Fogo Island; the dominant predator is coyote
(*Canis latrans*), which predate calves but it remains unclear what threat they
pose, if any, to adult caribou [@Rayl2014]. Given the minimal risk of predation to adults,
we propose that forage limitation is more likely to drive movement between
islands than predator avoidance. For several decades after their introduction to
Fogo Island, the caribou population did not exceed ~100 individuals
[@Bergerud1989, Newfoundland and Labrador Wildlife Division, unpublished data].
During the 1990s, population density reached ~300 animals and anecdotal evidence
suggests caribou were first seen on nearby islands during this same period.
Increased competition for resources on Fogo due to this increase in density
could have precipitated movement to other islands.

@@

Forage limitation may be the ultimate explanation for caribou swimming between
oceanic islands, but a number of proximate mechanisms remain untested. Our
observations of caribou swimming in the ocean lead to five questions for future
research:

@@

(ref:fig2) A) Map of the Fogo Island archipelago with swimming events between islands. B) Histogram displaying the frequency of swimming events throughout the year. Note, colours correspond to individual caribou and vertical black bars delineate the ice-free season (April 1 to December 31). C) Inset of swimming events between small islands on the northern coast of Fogo island. D) Inset of swimming events between islands on the southern coast of Fogo island. Note, the grey diamond represents the location we observed an adult male caribou swimming (see Figure 1).
